### 2016-03-11 paper/icfp16-submission/semantics.tex
state lemma that resolve(_, s) ~ resolve(a, t)
sketch induction proof, need to fill in cases in inductive step

\paragraph{Narrowing}
%
Only a successful call to \forcesym can change the partial
input type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$
(\ie $\trace'$ is a single-step extension of $\trace$)
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{\estuck'}{\vsu'}{\tsu'}$ invokes \forcesym.
\end{lem}

@@

\paragraph{Preservation}
We prove that each evaluation step \emph{refines} the partial input type
of $f$, \ie\ preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
$\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  First note that by Lemma~\ref{lem:force-inst} we can immediately
  discharge the \rulename{E-*-Bad} rules as they cannot change
  \ptype{\trace}{f} at all, and are thus trivially
  compatibility-preserving. For the \rulename{E-*-Good} rules we can
  show that, by virtue of \forcesym succeeding, all must preseve
  compatibility.
  % Note that all rules preserve partial types with the exception of when
  % \forcesym\ is called on a hole, in which case we may instantiate the hole with
  % a concrete value.
  % %
  % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
\end{proof}

\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
and $\trace'$ is a single-step extension of $\trace$
and $\resolve{\vhole{\thole}}{\vsu} \neq \resolve{\vhole{\thole}}{\vsu'}$
then $\vsu' = \vsu[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\vhole{\thole}$ does not change, so if the resolved values differ then
  $\vsu \neq \vsu'$.
  %
  Note that only \forcesym can update \vsu\
  and only the first case,
  when we extend \vsu\ with $[\ehole \mapsto v]$,
  does.
\end{proof}

@@

\begin{lem}
\label{lem:k-stuck}
For all $t$,
if $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$ and
   $\tincompat{t}{\ptype{\trace}{f}}$,
   then there exists a $v$ such that $\hastype{v}{t}$ and
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$ in at most
   $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  We can construct $v$ from $\trace$ as follows.
  %
  Let
  $$
  \trace_i = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},
             \ldots,
             \triple{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}},
             \triple{e_{i}}{\vsu_{i}}{\tsu_{i}}
  $$
  be the shortest prefix of $\trace$ such that
  $\tincompat{\ptype{\trace_i}{f}}{t}$.
  %
  We will show that $\resolve{\vhole{\thole}}{\vsu_{i-1}}$
  must contain some other hole $\ehole'$ which is instantiated with a
  concrete value at step $i$.
  %
  Furthermore, $\ehole'$ is instantiated in such a way that
  $\vincompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$.
  \ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
  %
  Finally, we will show that if we had instantiated $\ehole'$ with a
  value such that
  $\vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$, the current
  step would have gotten $\stuck$.

@@

  Next, we introduce a notion of \emph{value contexts}, defined
  analogously to evaluation contexts.
  %
  $$
    V ::= \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V}
  $$
  %
  Since $\vsu_{i-1}$ and $\vsu_{i}$ differ only in $\ehole'$ but the resolved
  values differ,
  $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \vcompat{\invctx{\ehole'[\thole']}}{t}$
  and
  $\resolve{\vhole{\thole}}{\vsu_{i}} = \vincompat{\invctx{v'}}{t}$.
  %
  Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
  we show by case analysis on the evaluation relation that
  $\invctx{u}$ could not make progress at this step, and thus
  $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{proof}

>>>>

\paragraph{Narrowing}
%
Only a successful call to \forcesym can change the partial
input type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
(\ie $\trace'$ is a single-step extension of $\trace$)
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{\tsu}{e'}{\vsu'}{\tsu'}$
 successfully invokes \forcesym.
%
\ES{not clear we still need this}
\end{lem}

@@

\paragraph{Preservation}
We prove that each evaluation step \emph{refines} the partial input type
of $f$, \ie\ preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$ and
$\trace'$ is a single-step extension of $\trace$, % \defeq \trace, \triple{\estuck'}{\vsu'}{\tsu'}$
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
%
\ES{not clear we still need this}
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  First note that by Lemma~\ref{lem:force-inst} we can immediately
  discharge the \rulename{E-*-Bad} rules as they cannot change
  \ptype{\trace}{f} at all, and are thus trivially
  compatibility-preserving. For the \rulename{E-*-Good} rules we can
  show that, by virtue of \forcesym succeeding, all must preseve
  compatibility.
  % Note that all rules preserve partial types with the exception of when
  % \forcesym\ is called on a hole, in which case we may instantiate the hole with
  % a concrete value.
  % %
  % But $\typeof{\vhole{\thole}} = \thole$, which is compatible with any type.
\end{proof}

\begin{lem}
\label{lem:resolve-compat}
For all traces
$\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots,\triple{e}{\vsu}{\tsu}$,
$\vcompat{\resolve{\ehole}{\vsu}}{\resolve{\thole}{\tsu}}$.
\end{lem}
\begin{proof}
  By induction on $\trace$.
  %
  In the base case $\trace = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}$
  and $\vhole{\thole}$ is trivially compatible with $\thole$.
  %
  In the inductive case, consider the single-step extension of $\trace$,
  $\trace' = \trace,\triple{e'}{\vsu'}{\tsu'}$.
  %
  We show by case analysis on the evaluation rules that if
  $\vcompat{\resolve{\ehole}{\vsu}}{\resolve{\thole}{\tsu}}$, then
  $\vcompat{\resolve{\ehole}{\vsu'}}{\resolve{\thole}{\tsu'}}$.
  %
  \ES{fill in cases in the appendix}
\end{proof}

\begin{lem}
\label{lem:vsu-ext}
If  $\trace \defeq \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},\ldots$
and $\trace'$ is a single-step extension of $\trace$
and $\resolve{\vhole{\thole}}{\vsu} \neq \resolve{\vhole{\thole}}{\vsu'}$
then $\vsu' = \vsu[\ehole_1 \mapsto v_1] \ldots [\ehole_n \mapsto v_n]$.
\end{lem}
\begin{proof}
  By case analysis on the evaluation rules.
  %
  $\vhole{\thole}$ does not change, so if the resolved values differ then
  $\vsu \neq \vsu'$.
  %
  Note that only \forcesym can update \vsu\
  and only the first case,
  when we extend \vsu\ with $[\ehole \mapsto v]$,
  does.
  %
  \ES{fill in cases in the appendix}
\end{proof}

@@

\begin{lem}
\label{lem:k-stuck}
For all $t$,
if $\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{e}{\vsu}{\tsu}$ and
   $\tincompat{t}{\ptype{\trace}{f}}$,
   then there exists a $v$ such that $\hastype{v}{t}$ and
   $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$ in at most
   $k$ steps, where $k$ is the length of $\trace$.
\end{lem}
\begin{proof}
  We can construct $v$ from $\trace$ as follows.
  %
  Let
  $$
  \trace_i = \triple{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu},
             \ldots,
             \triple{e_{i-1}}{\vsu_{i-1}}{\tsu_{i-1}},
             \triple{e_{i}}{\vsu_{i}}{\tsu_{i}}
  $$
  be the shortest prefix of $\trace$ such that
  $\tincompat{\ptype{\trace_i}{f}}{t}$.
  %
  We will show that $\resolve{\vhole{\thole}}{\vsu_{i-1}}$
  must contain some other hole $\ehole'$ which is instantiated with a
  concrete value at step $i$.
  %
  Furthermore, $\ehole'$ is instantiated in such a way that
  $\vincompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$.
  %
  Finally, we will show that if we had instantiated $\ehole'$ with a
  value such that
  $\vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{t}$, the current
  step would have gotten $\stuck$.

@@

  Next, we introduce a notion of \emph{value contexts}, defined
  analogously to evaluation contexts.
  %
  $$
    V ::= \bullet \spmid \vnode{t}{V}{v}{v} \spmid \vnode{t}{v}{V}{v} \spmid \vnode{t}{v}{v}{V}
  $$
  %
  Since $\vsu_{i-1}$ and $\vsu_{i}$ differ only in $\ehole'$ but the resolved
  values differ,
  $\resolve{\vhole{\thole}}{\vsu_{i-1}} = \invctx{\ehole'[\thole']}$
  and
  $\resolve{\vhole{\thole}}{\vsu_{i}} = \invctx{v'}$.
  %
  Furthermore, by Lemma~\ref{lem:resolve-compat} we know that
  $$
  \vcompat{\resolve{\vhole{\thole}}{\vsu_{i-1}}}{\resolve{\thole}{\tsu_{i-1}}}
  =
  \tcompat{\ptype{i-1}{f}}{t}
  $$
  $$
  \vcompat{\resolve{\vhole{\thole}}{\vsu_{i}}}{\resolve{\thole}{\tsu_{i}}}
  =
  \tincompat{\ptype{i}{f}}{t}
  $$
  %\ES{need to connect \ptype{\trace{i}}{f} and \resolve{\vhole{\thole}}{\vsu_i}}
  %
  Now, consider any value $u$ such that $\vcompat{\invctx{u}}{t}$,
  we show by case analysis on the evaluation relation that
  $\invctx{u}$ could not make progress at this step, and thus
  $\steps{\eapp{f}{\invctx{u}}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{proof}
